CF Nutrition Pre- and Post-CFTR Modulator

### 2022-10-04 Edith Zemanick/CF Nutrition/cf_nutrition.qmd
Update cf_nutrition.qmd
missing data report

```{r}
# Import all data
df = read.csv("/Users/timvigers/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis.csv",na.strings = "",check.names = F)
df$Timepoint = factor(df$Timepoint,levels = c("Pre","Post"))
```

@@

# Baseline value grouped by first modulator

```{r results='asis'}
t1 <- tableby(`First Modulator` ~ `Vitamin A` + `Vitamin D` + `Vitamin E Alpha` +
  `ppFEV1` + `BMI Percentile` +
  `ALT` + `High ALT (3x)` + `High ALT (5x)` +
  `AST` + `High AST (3x)` + `High AST (5x)` +
  `GGTP` + `High GGTP (3x)` + `High GGTP (5x)`,
data = df[df$Timepoint == "Pre", ]
)
summary(t1, pfootnote = T)
```

# Pre/post comparison

```{r results='asis'}
p <- paired(Timepoint ~ `Vitamin A` + signed.rank(`Vitamin D`) + `Vitamin E Alpha` +
  `ppFEV1` + `BMI Percentile` +
  signed.rank(`ALT`) + `High ALT (3x)` + notest(`High ALT (5x)`) +
  signed.rank(`AST`) + `High AST (3x)` + notest(`High AST (5x)`) +
  signed.rank(`GGTP`) + notest(`High GGTP (3x)`) + notest(`High GGTP (5x)`),
data = df, id = ID
)
summary(p, pfootnote = T)
```

>>>>

```{r warning=FALSE}
# Import datasets
vit <- read_excel("/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report Vitamins_throughDecember2021_revised20220712.xlsx")
mods <- read_excel("/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report CFTR Modulators_throughDecember2021.xlsx")
pfts <- read_excel("/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report_PFTs_ThroughDecember2021.xlsx")
lfts <- read_excel("/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Raw/Report_LFTs_ThroughDecember2021.xlsx")
geno <- read_excel("/Volumes/Documents/Work/Vigers/CF/CF Info/Sweat Tests from Internal Database_genotype class_withPancStatus.xlsx", na = ".")
# Format
vit$VitaminA_Retinol <- as.numeric(vit$VitaminA_Retinol)
vit$`Age at Test (years)` <- NULL
vit$Date <- as.Date(vit$Date)
vit$`Vitamin E Gamma` <- NULL
mods$`Earliest Date in Registry` <- as.Date(mods$`Earliest Date in Registry`)
pfts <- pfts %>%
  select(PatientID, Sex, Date, `Height (cm)`:`FEV1 % pred`) %>%
  rename(`Patient ID` = PatientID)
pfts$Date <- as.Date(pfts$Date)
lfts <- lfts %>% select(`Patient ID`, Date:GGTP)
lfts$Date <- as.Date(lfts$Date)
geno <- geno %>%
  select(PatientID, genoRisk, PancreaticStatus) %>%
  distinct() %>%
  rename(
    `Patient ID` = PatientID, `Genotype Risk` = genoRisk,
    `Pancreatic Status` = PancreaticStatus
  )
# Get first modulator and start date
mods <- mods %>%
  filter(Modulator %in% c("Kalydeco", "Trikafta")) %>%
  pivot_wider(
    id_cols = `Patient ID`, names_from = Modulator,
    values_from = c(`Earliest Date in Registry`, `Age at Start (years)`)
  )
mods$Start <- pmin(mods$`Earliest Date in Registry_Trikafta`,
  mods$`Earliest Date in Registry_Kalydeco`,
  na.rm = T
)
mods$tf <- mods$`Earliest Date in Registry_Trikafta` == mods$Start
mods$kf <- mods$`Earliest Date in Registry_Kalydeco` == mods$Start
mods <- mods %>% unite(First, tf, kf)
mods$`Age at Start` <- pmin(mods$`Age at Start (years)_Trikafta`,
  mods$`Age at Start (years)_Kalydeco`,
  na.rm = T
)
mods$First <- factor(mods$First,
  levels = c("FALSE_TRUE", "NA_TRUE", "TRUE_NA"),
  labels = c("Kalydeco", "Kalydeco", "Trikafta")
)
# Combine
vit <- full_join(vit, pfts, by = c("Patient ID", "Date"))
vit <- full_join(vit, lfts, by = c("Patient ID", "Date"))
vit <- left_join(vit, mods %>% select(`Patient ID`, Start, First, `Age at Start`),
  by = "Patient ID"
)
# Get pre vs. post info, time from start
vit$`Pre-/Post-` <- factor(vit$Date >= vit$Start,
  levels = c("FALSE", "TRUE"),
  labels = c("Pre", "Post")
)
vit$`Days from Start` <- as.numeric(difftime(vit$Date, vit$Start, units = "days"))
# Filter out those more than 3 years before start, and within 90 days after start
vit <- vit %>% filter(
  `Days from Start` > -3 * 365,
  `Days from Start` < 0 | `Days from Start` >= 90
)
# Order variables, fill sex
vit <- vit %>%
  select(
    `Patient ID`, Sex, Date, Start, First, `Age at Start`,
    `Pre-/Post-`, `Days from Start`, everything()
  ) %>%
  group_by(`Patient ID`) %>%
  fill(Sex, .direction = "downup")
# Average pre-mod values, get most recent vitamin values and and any other variables within 30 days
pre_ids <- vit %>%
  group_by(`Patient ID`) %>%
  filter(`Pre-/Post-` == "Pre", !is.na(VitaminA_Retinol)) %>%
  mutate(n = n()) %>%
  filter(n >= 2) %>%
  .$`Patient ID` %>%
  unique() # At least two pre measures
pre <- vit %>%
  filter(`Patient ID` %in% pre_ids) %>%
  summarise(
    `First Modulator` = as.character(First[1]), Timepoint = "Pre",
    `Age at Start` = `Age at Start`[1], Sex = Sex[1],
    Start = as.character(Start[1]), `Latest Visit` = NA,
    across(`Days from Start`:GGTP, ~ mean(.x, na.rm = TRUE))
  )
# Get visits within 30 days prior to latest vitamin measure post-CFTR modulator
post_cutoff <- vit %>%
  group_by(`Patient ID`) %>%
  filter(`Pre-/Post-` == "Post", !is.na(VitaminA_Retinol)) %>%
  mutate(latest = max(Date), date_cutoff = latest - lubridate::days(90)) %>%
  select(`Patient ID`, latest, date_cutoff) %>%
  distinct()
vit <- left_join(vit, post_cutoff, by = "Patient ID")
# Summarise visits into one row
post <- vit %>%
  filter(Date >= date_cutoff & Date <= latest) %>%
  arrange(`Patient ID`, Date) %>%
  summarise(
    `First Modulator` = as.character(First[1]), Timepoint = "Post",
    `Age at Start` = `Age at Start`[1], Sex = Sex[1],
    Start = as.character(Start[1]),
    `Latest Visit` = as.character(latest[1]),
    across(
      `Days from Start`:GGTP,
      ~ if (all(is.na(.))) NA else last(na.omit(.))
    )
  ) %>%
  rename()
df <- bind_rows(pre, post) %>%
  mutate_all(~ ifelse(is.nan(.), NA, .))
df$Timepoint <- factor(df$Timepoint, levels = c("Pre", "Post"))
# Pancreatic status and LFTs
df$`Elevated LFTs (2X)` <- factor((!is.na(df$ALT) & df$ALT > 2 * 30) |
  !is.na(df$AST) & df$AST > 2 * 30 |
  !is.na(df$GGTP) & df$GGTP > 2 * 20,
levels = c("TRUE", "FALSE"), labels = c("Yes", "No")
)
df$`Elevated LFTs (2X)`[rowSums(is.na(df[, c("ALT", "AST", "GGTP")])) == 3] <- NA
df$`Elevated LFTs (3X)` <- factor((!is.na(df$ALT) & df$ALT > 3 * 30) |
  !is.na(df$AST) & df$AST > 3 * 30 |
  !is.na(df$GGTP) & df$GGTP > 3 * 20,
levels = c("TRUE", "FALSE"), labels = c("Yes", "No")
)
df$`Elevated LFTs (3X)`[rowSums(is.na(df[, c("ALT", "AST", "GGTP")])) == 3] <- NA
df$`Elevated LFTs (5X)` <- factor((!is.na(df$ALT) & df$ALT > 5 * 30) |
  !is.na(df$AST) & df$AST > 5 * 30 |
  !is.na(df$GGTP) & df$GGTP > 5 * 20,
levels = c("TRUE", "FALSE"), labels = c("Yes", "No")
)
df$`Elevated LFTs (5X)`[rowSums(is.na(df[, c("ALT", "AST", "GGTP")])) == 3] <- NA
# Add genetic info
df <- left_join(df, geno, by = "Patient ID")
# Delete those without paired obs.
df <- df %>%
  group_by(`Patient ID`) %>%
  filter(n() == 2) %>%
  arrange(`Patient ID`, Timepoint)
# Write data so Edith can check
write.csv(df, file = "/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_data.csv", row.names = F, na = "")
write.csv(df[df$`First Modulator` == "Trikafta", ], file = "/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/analysis_data_eti_only.csv", row.names = F, na = "")
# Write missing
mrns <- read_excel("/Volumes/Documents/Work/Vigers/CF/CF Info/All Patients with CF or Non-CF_MRNs.xlsx")

miss_post_pft <- df %>%
  filter(Timepoint == "Post", is.na(`FEV1 % pred`)) %>%
  mutate(
    MRN = mrns$MRN[match(`Patient ID`, mrns$PatientID)],
    `Start Plus 90` = parse_date(Start, approx = F) + lubridate::days(90),
    `Latest Minus 30` = parse_date(`Latest Visit`, approx = F) - lubridate::days(30)
  ) %>%
  select(`Patient ID`, MRN, `First Modulator`, Start, `Start Plus 90`, `Latest Visit`, `Latest Minus 30`)

write.csv(miss_post_pft, file = "/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/missing_post_pft.csv", row.names = F, na = "")

miss_post_lft <- df %>%
  filter(Timepoint == "Post", is.na(ALT)) %>%
  mutate(
    MRN = mrns$MRN[match(`Patient ID`, mrns$PatientID)],
    `Start Plus 90` = parse_date(Start, approx = F) + lubridate::days(90),
    `Latest Minus 30` = parse_date(`Latest Visit`, approx = F) - lubridate::days(30)
  ) %>%
  select(`Patient ID`, MRN, `First Modulator`, Start, `Start Plus 90`, `Latest Visit`, `Latest Minus 30`)

write.csv(miss_post_lft, file = "/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/missing_post_lft.csv", row.names = F, na = "")

miss_pre_pft <- df %>%
  filter(Timepoint == "Pre", is.na(`FEV1 % pred`)) %>%
  mutate(
    MRN = mrns$MRN[match(`Patient ID`, mrns$PatientID)]
  ) %>%
  select(`Patient ID`, MRN, `First Modulator`, Start)

write.csv(miss_pre_pft, file = "/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/missing_pre_pft.csv", row.names = F, na = "")

miss_pre_lft <- df %>%
  filter(Timepoint == "Pre", is.na(ALT)) %>%
  mutate(
    MRN = mrns$MRN[match(`Patient ID`, mrns$PatientID)]
  ) %>%
  select(`Patient ID`, MRN, `First Modulator`, Start)

write.csv(miss_pre_lft, file = "/Volumes/Documents/Work/Vigers/CF/Edith Zemanick/CF Nutrition/Data_Cleaned/missing_pre_lft.csv", row.names = F, na = "")
```

@@

-   Elevated liver labs were considered at two levels:

    -   ALT, AST, or GGTP 3 times the normal level

    -   ALT, AST, or GGTP 5 times the normal level

-   Normal liver labs were considered:

    -   ALT: 30 U/L

    -   AST: 30 U/L

    -   GGTP: 20 U/L

# Ivacaftor and ETI

## Baseline value grouped by first modulator

```{r results='asis'}
t1 <- tableby(`First Modulator` ~ `Age at Start` + Sex + `ALT` + `AST` + `GGTP` +`Elevated LFTs (2X)`+ `Elevated LFTs (3X)` + `Elevated LFTs (5X)`+`Pancreatic Status`,data = df[df$Timepoint == "Pre", ],numeric.stats = c("Nmiss","medianrange"),numeric.test="kwt")
summary(t1, pfootnote = T)
```

## Pre/post comparison

```{r results='asis'}
p <- paired(Timepoint ~ `VitaminA_Retinol` + `25OH-Vitamin D` + `Vitamin E Alpha` + `FEV1 % pred` + `BMI percentile (CDC)` + `ALT` + `AST` + `GGTP` + `Elevated LFTs (2X)` + `Elevated LFTs (3X)`+`Elevated LFTs (5X)`,data = df, id = `Patient ID`,numeric.stats = c("Nmiss","medianrange"),numeric.test="signed.rank",signed.rank.exact = FALSE)
summary(p, pfootnote = T)
```

## Pre/post comparison by pancreatic status

One participant with unknown pancreatic status was removed in order to calculate p values.

```{r results='asis'}
panc_df = df %>% filter(`Pancreatic Status`!= "Unknown")

p <- paired(Timepoint ~ `VitaminA_Retinol` + `25OH-Vitamin D` + `Vitamin E Alpha` + `FEV1 % pred` + `BMI percentile (CDC)` + `ALT` + `AST` + `Elevated LFTs (3X)` + `Elevated LFTs (5X)`,data = panc_df, id = `Patient ID`,strata = `Pancreatic Status`,numeric.stats = c("Nmiss","medianrange"),numeric.test="signed.rank",signed.rank.exact = FALSE)
summary(p, pfootnote = T)
```

# ETI only

## Pre/post comparison

```{r results='asis'}
p <- paired(Timepoint ~ `VitaminA_Retinol` + `25OH-Vitamin D` + `Vitamin E Alpha` + `FEV1 % pred` + `BMI percentile (CDC)` + `ALT` + `AST` + `GGTP` + `Elevated LFTs (3X)` + `Elevated LFTs (5X)` + `Pancreatic Status`,data = df[df$`First Modulator`=="Trikafta",], id = `Patient ID`,numeric.stats = c("Nmiss","medianrange"),numeric.test="signed.rank",signed.rank.exact = FALSE)
summary(p, pfootnote = T)
```
